---
title: "RCN DOC Exports - Predictive Modeling"
author: "Gavin McNicol"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: monochrome
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

Code for predictive modeling of stream dissolved organic carbon (DOC) using random forest and spatial (forward) feature selection in support of the manuscript:

*Carbon Export along Coastal Temperate Rainforest Margins Dominated by Small Watersheds* by G. McNicol, E. Hood, D. E. Butman, S. Tank, I.J.W. Giesbrecht, W. Floyd, D. D’Amore, J.B. Fellman (2022-2023)

## Analysis

```{r load-packages, include = F}
source("code/libraries.R") # loads packages as pkg_req
lapply(pkg_req, require, character.only = TRUE)
if(pkg_req %in% (.packages())){
  pkg_req_true = "Yes"
} else {
  pkg_req_true = "No"
}
```

Packages loaded: `r pkg_req_true`

```{r load-mytheme, include = F}
source("code/mytheme.R")  # loads ggplot theme
```

## Part 1: Create Training Data

### Load and View DOC Data

Much of Part 1 code can be suppressed `eval = F` for Part 1 as `training-data.csv` exists in repo.

```{r load-doc, echo = F, message = F, eval = F}
doc <- read_csv("data/annual-doc-yield.csv") %>% 
  dplyr::select(IP_ID, Use, DOC_yield = annual_DOC_yield, DOC_yield_sd = sd_annual_DOC_yield)
glimpse(doc)
```


### Load and View Predictive Features

Exclude geolocation features: `Poly_x`, `Poly_y`

```{r load-predictors, echo = F, message = F}
feat <- read_csv("data/WtsData_For_RCN_DOCflux.csv") %>% 
  dplyr::select(-"Poly_x", -"Poly_y")
glimpse(feat)
```

## Load in LAT/LONs

```{r load-lat-lon, echo = F, message = F}
feat_lat_lon <- read_csv("data/Watersheds_of_the_northern_Pacific_coastal_temperate_rainforest_margin_Centroids_WGS84.csv") %>% 
  dplyr::select(IP_ID, "lat" = Lat, "lon" = Lon) %>% 
  right_join(feat, by = "IP_ID")
```


### Feature Objects

```{r subset-features, include = F}
# combine features 
feat_names <- feat %>% 
  dplyr::select(-IP_ID, -Cluster) %>% 
  names()
feat_l <- length(feat_names)
```

`feat_l` = `r feat_l`  

`feat_names`: `r cbind(feat_names)`

### Geneate Combinations

Minimum number of features = 2
Maximum number of features = 5

Takes a couple of minute to run for max. = 5. Longer for more.

```{r get-feat-combin, include = F, eval = F}
feat_index <- expand.grid(0:1, 0:1, 0:1, 0:1, 0:1,
            0:1, 0:1, 0:1, 0:1, 0:1,
            0:1, 0:1, 0:1, 0:1, 0:1,
            0:1, 0:1) %>% 
  as_tibble() %>% 
  rowwise() %>% 
  mutate(sum = sum(c_across(Var1:Var17))) %>% 
  filter(sum %in% c(2, 3, 4, 5)) %>% 
  dplyr::select(-sum)
```

```{r get-feat-combin-2, include = F, eval = F}
feat_comb_list <- list()
for(i in 1:nrow(feat_index)){
  feat_comb_list[[i]] <- feat_names[which(feat_index[i,] == 1)]
}
feat_comb <- length(feat_comb_list)
```

**Output:** Predictive modeling feature combinations list, `data/feat_comb_list.rds`

```{r output-feat-comb, eval = F}
saveRDS(feat_comb_list, "data/feat_comb_list.rds")
```

There are ~9,000 feature combinations for 2-5 predictors.

### Join DOC and Features

```{r join-data, include = F, eval = F}
data <- doc %>% 
  left_join(feat_lat_lon, by = "IP_ID")
```

**Output:** Predictive modeling training dataset, `data/training-data.csv`

```{r output-training-data, eval = F}
write_csv(data, "data/training-data.csv")
```

## Part 2: Train Random Forest

```{r read-data, output = F, message = F}
train <- read_csv("data/training-data.csv")
feat_comb_list <- readRDS("data/feat_comb_list.rds")
```

```{r  set-seed, output = F, message = F}
set.seed(23)
IP_IDs <- train %>% 
  select(IP_ID) %>% pull
```

### Cluster Watersheds

```{r cluster-watersheds}
x <- train$lon
y <- train$lat
xy <- SpatialPointsDataFrame(matrix(c(x,y), ncol=2), data.frame(IP_ID=train$IP_ID),
                             proj4string = CRS("+proj=longlat +ellps=WGS84 +datum=WGS84"))
plot(xy)

# calculate Euclidian distance
site_dist <- xy %>% distm()

# cluster
site_clusters <- hclust(as.dist(site_dist), method ='complete')

# define 1000 km threshold
d <- 100000
xy$fold <- cutree(site_clusters, h=d)
xy <- xy %>% as_tibble() %>%
  dplyr::select(IP_ID, fold)

xy %>% 
  count(fold)

# rejoin folds
train <- train %>%
  left_join(xy, by = c("IP_ID")) %>%
  dplyr::select(fold = fold, everything())


# check train length and fold number
train_l <- length(train$lat)
folds <- max(train$fold)
watersheds <- length(unique(as.character(train$IP_ID)))

```


## Start FFS
```{r set-up-folds, echo = F, message = F}

train_feat <- list()
train_label <- list()

for (i in 1:folds) {
  train_label[[i]] <- train %>% filter(!fold == i) %>% dplyr::select(DOC_yield) %>% pull()
  train_feat[[i]] <- train %>% filter(!fold == i) %>% dplyr::select(all_of(feat_names)) %>% as_tibble()
}

folds_index <- list() # need list within list (each inner list has all fold site except LOSO fold)
for (i in 1:folds) {
  x <- list()
  folds_index[[i]] <- x
}

for (i in 1:folds) {
  for (j in 1:folds) {
    folds_index[[i]][[j]] <- train %>% 
      filter(!fold == i) %>% 
      mutate(index = 1:n()) %>% 
      filter(!fold == j) %>% 
      dplyr::select(index) %>% 
      pull()
  }
  folds_index[[i]] <- folds_index[[i]][-i]
}

```

###  FFS by training and evaluating models

```{r ffs, echo = F, message = F, warning = F}

## Train multiple RFs (with CV) for FFS to select the first pair
# outer loop lists
rf_metrics <- list()

for(j in 1677:length(feat_comb_list)) {
  # set up RF lists
  tgrid <- list()
  myControl <- list()
  rf_model <- list()
  rf_varimp <- list()
  
  for (i in 1:folds) {
    
    num.trees.rf <- ifelse(length(feat_comb_list[[j]]) < 5, 50, 100)
    
    ## Create tune-grid
    tgrid <- expand.grid(
      mtry = c(2),
      splitrule = "variance", 
      min.node.size = c(50)
    )
    
    ## Create trainControl object
    myControl <- trainControl(
      method = "none",
      classProbs = FALSE,
      allowParallel = TRUE
    )
    
    ## train rf on folds
    rf_model[[i]] <- train(
      x = train_feat[[i]][,feat_comb_list[[j]]], 
      y = train_label[[i]],
      num.trees = num.trees.rf, # start 300 drop to 100
      method = 'ranger',
      trControl = myControl,
      tuneGrid = tgrid,
      metric = "MAE"
    )
    # print(i)
  }
 
  # get all hold out predictions
  rf.pred <- list()
  for (i in 1:folds) {
    rf.pred[[i]] <- train %>% 
      filter(fold == i) %>%   
      mutate(DOC_yieldP = predict(rf_model[[i]], .),
             index = 1:n()) 
  }
  rf.pred.all <- bind_rows(rf.pred)
  
  rf_metrics[[j]] <- ungroup(rf.pred.all) %>% 
    # filter(!ID %in% c("USOWC")) %>%
    summarize(samples = n(),
              R2 = cor(DOC_yieldP, DOC_yield)^2,
              NSE = 1 - sum((DOC_yield - DOC_yieldP)^2) / sum((DOC_yield - mean(DOC_yield))^2),
              MAE = sum(abs(DOC_yield - DOC_yieldP))/n(),
              nMAE = MAE/sd(DOC_yield),
              MeanO = mean(DOC_yield),
              MedO = median(DOC_yield),
              sdO = sd(DOC_yield),
              MeanP = mean(DOC_yieldP),
              MedP = median(DOC_yieldP),
              sdP = sd(DOC_yieldP),
              nSD = sdP/sdO,
              Bias = mean(DOC_yieldP - DOC_yield),
              cBias = abs(Bias)/sum(abs(Bias))*100,
              predictors = list(feat_comb_list[[j]])) 
  
  print(j)
}

```
## Combine cross-validation metrics and write out data
```{r}

rf_metrics_all <- bind_rows(rf_metrics)  # first run, 1-1676
rf_metrics_all <- rf_metrics_all %>% 
  mutate(index = 1:n())

rf_metrics_all %>% 
  arrange(desc(NSE)) %>% 
  pull(predictors[[1]])


```


## Plot and save metrics

```{r}
# look at metrics distribution
rf_metrics_all %>% 
  ggplot(aes(R2)) +
  geom_histogram() + mytheme
ggsave(paste("output/ffs_r2.png", sep = ""),
       width = 8, height = 15, units = c("cm"), dpi = 300) 

rf_metrics_all %>% 
  ggplot(aes(MAE)) +
  geom_histogram() + mytheme
ggsave(paste("output/ffs_mae.png", sep = ""),
       width = 8, height = 15, units = c("cm"), dpi = 300) 


saveRDS(rf_metrics_all, "output/fs_all.rds")

rf_metrics_all <- readRDS("output/fs_all.rds")

```


### Train with feature subset for cross validation performance
```{r cv-performance, echo = F, message = F, warning = F}
# short vector
rf_metrics_all %>% 
  filter(index == 3814) %>% 
  pull(predictors)
  
feat_names <- c("PrctSnow", "Eref_avg")

# long vector (5 predictors)
# feat_names <- c("Elev_avg", "PAS_avg",  "MAT_avg", "TD_avg", "Eref_avg")

train_feat <- list()
train_label <- list()

for (i in 1:folds) {
  train_label[[i]] <- train %>% filter(!fold == i) %>% dplyr::select(DOC_yield) %>% pull()
  train_feat[[i]] <- train %>% filter(!fold == i) %>% dplyr::select(all_of(feat_names)) %>% as_tibble()
}

folds_index <- list() # need list within list (each inner list has all fold site except LOSO fold)
for (i in 1:folds) {
  x <- list()
  folds_index[[i]] <- x
}

for (i in 1:folds) {
  for (j in 1:folds) {
    folds_index[[i]][[j]] <- train %>% 
      filter(!fold == i) %>% 
      mutate(index = 1:n()) %>% 
      filter(!fold == j) %>% 
      dplyr::select(index) %>% 
      pull()
  }
  folds_index[[i]] <- folds_index[[i]][-i]
}

# set up RF lists
tgrid <- list()
myControl <- list()
rf_model <- list()
rf_varimp <- list()

for (i in 1:folds) {
  
  ## Create tune-grid
  tgrid <- expand.grid(
    mtry = c(2),
    splitrule = "variance", 
    min.node.size = c(2,6,12,18)
  )
  
  ## Create trainControl object
  myControl <- trainControl(
    method = "cv",
    classProbs = FALSE,
    allowParallel = TRUE,
    verboseIter = TRUE, 
    savePredictions = TRUE,
    index = folds_index[[i]]
  )
  
  ## train rf on folds
  rf_model[[i]] <- train(
    x = train_feat[[i]], 
    y = train_label[[i]],
    num.trees = 50, 
    method = 'ranger',
    # weights = train_label[[i]]/max(train_label[[i]])^2,
    trControl = myControl,
    tuneGrid = tgrid,
    importance = 'permutation',
    metric = "MAE"
  )
  print(i)
}

saveRDS(rf_model, "output/cv_rfs.rds")

```

## Get variable importance

```{r}

rf_model <- readRDS("output/cv_rfs.rds")
folds <- 34

## look at variable importance, create table of all site rankings
variable.imp <- list()
var.imp.ranks <- list()
variable.imp.single <- list()
for (i in 1:folds) {
  variable.imp[[i]] <- varImp(rf_model[[i]], scale = FALSE)
}
var.imp.names <- rownames(variable.imp[[1]]$importance)
for (i in 1:folds) {
  var.imp.ranks[[i]] <- variable.imp[[i]]$importance$Overall
  variable.imp.single[[i]] <- cbind(var.imp.names, var.imp.ranks[[i]])
  variable.imp.single[[i]] <- variable.imp.single[[i]] %>% as_tibble() %>% mutate(V2 = as.integer(V2))
  variable.imp.single[[i]] <- variable.imp.single[[i]] %>% arrange(desc(V2)) %>% dplyr::select(var.imp.names,V2)
}

for (i in 1:folds) {
  names(variable.imp.single[[i]]) <- c("Var",paste("Imp",i,sep=""))
  variable.imp.single[[i]] <- variable.imp.single[[i]] %>% arrange(Var)
}

# get full table of rf pred impotance
variable.importance <- as_tibble(bind_cols(variable.imp.single)) %>% 
  dplyr::select(1,c(seq(2,62,2))) %>% group_by(Var...1) %>% 
  mutate(AvgImp = mean(c(Imp1,Imp2,Imp3,Imp4,Imp5,Imp6,Imp7,Imp8,Imp9,Imp10,Imp11,Imp12,
                         Imp13,Imp14,Imp15,Imp16,Imp17,Imp18,Imp19,Imp20,Imp21,Imp22,
                         Imp23,Imp24,Imp25,Imp26,Imp27,Imp28,Imp29,Imp30,Imp31))) %>% 
  ungroup() %>% 
  mutate(scalingfactor = 100/max(AvgImp),
         AvgImpScaled = AvgImp*scalingfactor) %>% 
  arrange(desc(AvgImpScaled)) %>% 
  dplyr::select(-AvgImp, -scalingfactor)
view(variable.importance)

```

## Get CV scores

```{r}

# get all hold out predictions
rf.pred <- list()
for (i in 1:folds) {
  rf.pred[[i]] <- train %>% 
    filter(fold == i) %>%   
    mutate(DOC_yieldP = predict(rf_model[[i]], .),
           index = 1:n()) 
}
rf.pred.all <- bind_rows(rf.pred)

rf.pred.all %>% 
  ggplot(aes(DOC_yield, DOC_yieldP)) +
  geom_point(alpha = 0.5) + 
  scale_x_continuous(limits = c(0, 40)) +
  scale_y_continuous(limits = c(0, 40)) +
  geom_abline(slope = 1, intercept = 0) +
  # geom_smooth() +
  mytheme

ungroup(rf.pred.all) %>% 
  # filter(!ID %in% c("USOWC")) %>%
  summarize(samples = n(),
            R2 = cor(DOC_yieldP, DOC_yield)^2,
            NSE = 1 - sum((DOC_yield - DOC_yieldP)^2) / sum((DOC_yield - mean(DOC_yield))^2),
            MAE = sum(abs(DOC_yield - DOC_yieldP))/n(),
            nMAE = MAE/sd(DOC_yield),
            MeanO = mean(DOC_yield),
            MedO = median(DOC_yield),
            sdO = sd(DOC_yield),
            MeanP = mean(DOC_yieldP),
            MedP = median(DOC_yieldP),
            sdP = sd(DOC_yieldP),
            nSD = sdP/sdO,
            Bias = mean(DOC_yieldP - DOC_yield),
            cBias = abs(Bias)/sum(abs(Bias))*100)


# try bias correction
splinemod <- smooth.spline(y=rf.pred.all$DOC_yield[which(!is.na(rf.pred.all$DOC_yield))], x=rf.pred.all$DOC_yieldP[which(!is.na(rf.pred.all$DOC_yieldP))], spar = 2)
plot(rf.pred.all$DOC_yieldP,rf.pred.all$DOC_yield, xlim = c(0,40), ylim = c(0,40))
abline(0,1,col = 'grey')
lines(splinemod, col="black", lwd=5)

saveRDS(splinemod, "output/rf_bias_spline.rds")

```

## get bias corrected predictions
```{r}
DOC_yieldP_biascor <- predict(splinemod, rf.pred.all$DOC_yieldP)  
DOC_yieldP_biascor <- DOC_yieldP_biascor$y
rf.pred.all <- cbind(rf.pred.all, DOC_yieldP_biascor)

rf.pred.all %>% 
  ggplot(aes(DOC_yield, DOC_yieldP_biascor)) +
  geom_abline(slope = 1, intercept = 0, color = "grey") +
  geom_point(size = 2, alpha = 0.9) + 
  scale_x_continuous(limits = c(0, 40)) +
  scale_y_continuous(limits = c(0, 40)) +
  # geom_smooth() +
  labs(x = expression("Observed Yield (gC m"^{-2}*" y"^{-1}*")"), y = expression("Predicted Yield (Bias Cor.) (gC m"^{-2}*" y"^{-1}*")")) +
  mytheme
ggsave(paste("output/extrapolation_pred_vs_obs.png", sep = ""),
       width = 12, height = 12, units = c("cm"), dpi = 300)


ungroup(rf.pred.all) %>% 
  # filter(!ID %in% c("USOWC")) %>%
  summarize(samples = n(),
            R2 = cor(DOC_yieldP_biascor, DOC_yield)^2,
            NSE = 1 - sum((DOC_yield - DOC_yieldP_biascor)^2) / sum((DOC_yield - mean(DOC_yield))^2),
            MAE = sum(abs(DOC_yield - DOC_yieldP_biascor))/n(),
            nMAE = MAE/sd(DOC_yield),
            RMSE = sqrt(sum((DOC_yield - DOC_yieldP_biascor)^2)/n()),
            MeanO = mean(DOC_yield),
            MedO = median(DOC_yield),
            sdO = sd(DOC_yield),
            MeanP = mean(DOC_yieldP_biascor),
            MedP = median(DOC_yieldP_biascor),
            sdP = sd(DOC_yieldP_biascor),
            nSD = sdP/sdO,
            Bias = mean(DOC_yieldP_biascor - DOC_yield),
            cBias = abs(Bias)/sum(abs(Bias))*100) %>% view()

write.csv(rf.pred.all, "output/rf_pred_all.csv")

```


## Final RF model for predictions

```{r}

# set up folds
train_label <- train  %>% dplyr::select(DOC_yield) %>% pull()
train_feat <- train %>% dplyr::select(all_of(feat_names))

folds_index <- list() 
for (i in 1:folds) {
    folds_index[[i]] <- train %>% 
      mutate(index = 1:n()) %>% 
      filter(!fold == i) %>% 
      dplyr::select(index) %>% 
      pull()
}

## set up lists 
tgrid <- list()
myControl <- list()

## Create tune-grid
tgrid <- expand.grid(
  mtry = c(2),
  splitrule = "variance", 
  min.node.size = c(2)
)

## Create trainControl object
myControl <- trainControl(
    method = "cv",
    classProbs = FALSE,
    allowParallel = TRUE,
    verboseIter = TRUE, 
    savePredictions = TRUE,
    index = folds_index
  )

  ## train rf on folds
rf_model <- train(
    x = train_feat, 
    y = train_label,
    num.trees = 50, # start 300 
    method = 'ranger',
    # weights = train_label[[i]]/max(train_label[[i]])^2,
    trControl = myControl,
    tuneGrid = tgrid,
    importance = 'permutation',
    metric = "MAE"
  )

saveRDS(rf_model, "output/rf_final.rds")

```


## extrapolate predictions

```{r}
rf_model <- readRDS("output/rf_final.rds")

rcn.pred <- feat %>% 
  left_join(doc, by = c("IP_ID")) %>% 
    mutate(DOC_yieldP = predict(rf_model, .),
           index = 1:n()) 

rcn.pred %>% 
  ggplot(aes(DOC_yield, DOC_yieldP)) +
  geom_point(alpha = 0.5) + 
  scale_x_continuous(limits = c(0, 40)) +
  scale_y_continuous(limits = c(0, 40)) +
  geom_abline(slope = 1, intercept = 0) +
  geom_smooth() +
  mytheme

```

## Final Model Bias Cor

```{r}

# try bias correction
splinemod <- smooth.spline(y=rcn.pred$DOC_yield[which(!is.na(rcn.pred$DOC_yield))], x=rcn.pred$DOC_yieldP[which(!is.na(rcn.pred$DOC_yield))], spar = 2)
plot(rcn.pred$DOC_yieldP,rcn.pred$DOC_yield, xlim = c(0,40), ylim = c(0,40))
abline(0,1,col = 'grey')
lines(splinemod, col="black", lwd=5)

# get bias corrected predictions
DOC_yieldP_biascor <- predict(splinemod, rcn.pred$DOC_yieldP)  
DOC_yieldP_biascor <- DOC_yieldP_biascor$y
rcn.pred <- cbind(rcn.pred, DOC_yieldP_biascor)

# quick evaluation
rcn.pred %>% 
  ggplot(aes(DOC_yield, DOC_yieldP_biascor)) +
  geom_abline(intecept = 0, slope = 1, color = 'grey') +
  geom_point(size = 2) +
  scale_x_continuous(limits = c(0, 40)) +
  scale_y_continuous(limits = c(0, 40)) +
  # stat_smooth() +
  labs(x = expression("Observed Yield (gC m"^{-2}*" y"^{-1}*")"), y = expression("Predicted (Bias Cor.) Yield (gC m"^{-2}*" y"^{-1}*")")) +
  mytheme
ggsave(paste("output/pred_vs_obs_bias_cor.png", sep = ""),
       width = 12, height = 12, units = c("cm"), dpi = 300)

ungroup(rcn.pred) %>% 
  filter(!is.na(DOC_yield)) %>%
  summarize(samples = n(),
            R2 = cor(DOC_yieldP_biascor, DOC_yield)^2,
            NSE = 1 - sum((DOC_yield - DOC_yieldP_biascor)^2) / sum((DOC_yield - mean(DOC_yield))^2),
            MAE = sum(abs(DOC_yield - DOC_yieldP_biascor))/n(),
            nMAE = MAE/sd(DOC_yield),
            MeanO = mean(DOC_yield),
            MedO = median(DOC_yield),
            sdO = sd(DOC_yield),
            MeanP = mean(DOC_yieldP_biascor),
            MedP = median(DOC_yieldP_biascor),
            sdP = sd(DOC_yieldP_biascor),
            nSD = sdP/sdO,
            Bias = mean(DOC_yieldP_biascor - DOC_yield),
            cBias = abs(Bias)/sum(abs(Bias))*100) %>% view()

write.csv(rcn.pred, "output/rcn_yield.csv",
          row.names = FALSE)
```

## Final Prediction Evaluation

```{r}

rcn.pred <- read_csv("output/rcn_yield.csv")

summary(rcn.pred$DOC_yieldP_biascor)

rcn.pred %>% 
  arrange(desc(Area_km2)) %>% 
  filter(Area_km2 < 9000) %>% 
  mutate(Load_Tgy = DOC_yieldP_biascor*Area_km2*10^6*10^-12) %>% 
  summarize(Area_km2 = sum(Area_km2),
            RCN_Load_Tgy = sum(Load_Tgy)) %>% 
  mutate(RCN_DOC_yield_biascor = RCN_Load_Tgy*10^12*(1/Area_km2)*(1/10^6))

# uncertainty MAE (2.7295)
rcn.pred %>% 
  arrange(desc(Area_km2)) %>% 
  filter(Area_km2 < 9000) %>% 
  mutate(Load_Tgy = 2.73*Area_km2*10^6*10^-12) %>% 
  summarize(Area_km2 = sum(Area_km2),
            RCN_Load_Tgy = sum(Load_Tgy)) %>% 
  mutate(RCN_DOC_yield_biascor = RCN_Load_Tgy*10^12*(1/Area_km2)*(1/10^6))

    # uncertainty RMSE (4.35217)
rcn.pred %>% 
  arrange(desc(Area_km2)) %>% 
  filter(Area_km2 < 9000) %>% 
  mutate(Load_Tgy = 4.35*Area_km2*10^6*10^-12) %>% 
  summarize(Area_km2 = sum(Area_km2),
            RCN_Load_Tgy = sum(Load_Tgy)) %>% 
  mutate(RCN_DOC_yield_biascor = RCN_Load_Tgy*10^12*(1/Area_km2)*(1/10^6))

rcn.pred %>% 
  ggplot(aes(MAP_avg, DOC_yieldP_biascor)) +
  geom_point() +
  # scale_x_continuous(limits = c(0, 40)) +
  # scale_y_continuous(limits = c(0, 40)) +
  stat_smooth() +
  # geom_abline(intecept = 0, slope = 1) +
  mytheme

```



