---
title: "RCN DOC Exports - Predictive Modeling"
author: "Gavin McNicol"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: monochrome
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

Code for predictive modeling of stream dissolved organic carbon (DOC) using random forest and spatial (forward) feature selection in support of the manuscript:

*Carbon Export along Coastal Temperate Rainforest Margins Dominated by Small Watersheds* by G. McNicol, E. Hood, D. E. Butman, S. Tank, I.J.W. Giesbrecht, W. Floyd, D. D’Amore, J.B. Fellman (2022-2023)

## Analysis

```{r load-packages, include = F}
source("code/libraries.R") # loads packages as pkg_req
lapply(pkg_req, require, character.only = TRUE)
if(pkg_req %in% (.packages())){
  pkg_req_true = "Yes"
} else{
  pkg_req_true = "No"
}
```

Packages loaded: `r pkg_req_true`

```{r load-mytheme, include = F}
source("code/mytheme.R")  # loads ggplot theme
```

### Load DOC and Predictor Data





