---
title: "RCN DOC Exports - Predictive Modeling"
author: "Gavin McNicol"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: monochrome
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

Code for predictive modeling of stream dissolved organic carbon (DOC) using random forest and spatial (forward) feature selection in support of the manuscript:

*Carbon Export along Coastal Temperate Rainforest Margins Dominated by Small Watersheds* by G. McNicol, E. Hood, D. E. Butman, S. Tank, I.J.W. Giesbrecht, W. Floyd, D. D’Amore, J.B. Fellman (2022-2023)

## Analysis

```{r load-packages, include = F}
source("code/libraries.R") # loads packages as pkg_req
lapply(pkg_req, require, character.only = TRUE)
if(pkg_req %in% (.packages())){
  pkg_req_true = "Yes"
} else{
  pkg_req_true = "No"
}
```

Packages loaded: `r pkg_req_true`

```{r load-mytheme, include = F}
source("code/mytheme.R")  # loads ggplot theme
```

## Part 1: Create Training Data

### Load and View DOC Data

Much of Part 1 code can be suppressed `eval = F` for Part 1 as `training-data.csv` exists in repo.

```{r load-doc, echo = F, message = F, eval = F}
doc <- read_csv("data/annual-doc-yield.csv") %>% 
  dplyr::select(IP_ID, Use, DOC_yield = annual_DOC_yield, DOC_yield_sd = sd_annual_DOC_yield)
glimpse(doc)
```


### Load and View Predictive Features

Exclude geolocation features: `Poly_x`, `Poly_y`

```{r load-predictors, echo = F, message = F}
feat <- read_csv("data/WtsData_For_RCN_DOCflux.csv") %>% 
  dplyr::select(-"Poly_x", -"Poly_y")
glimpse(feat)
```

### Feature Objects

```{r subset-features, include = F}
# combine features 
feat_names <- feat %>% 
  dplyr::select(-IP_ID, -Cluster) %>% 
  names()
feat_l <- length(feat_names)
```

`feat_l` = `r feat_l`  

`feat_names`: `r cbind(feat_names)`

### Geneate Combinations

Minimum number of features = 2
Maximum number of features = 5

Takes a couple of minute to run for max. = 5. Longer for more.

```{r get-feat-combin, include = F, eval = F}
feat_index <- expand.grid(0:1, 0:1, 0:1, 0:1, 0:1,
            0:1, 0:1, 0:1, 0:1, 0:1,
            0:1, 0:1, 0:1, 0:1, 0:1,
            0:1, 0:1) %>% 
  as_tibble() %>% 
  rowwise() %>% 
  mutate(sum = sum(c_across(Var1:Var17))) %>% 
  filter(sum %in% c(2, 3, 4, 5)) %>% 
  dplyr::select(-sum)
```

```{r get-feat-combin-2, include = F, eval = F}
feat_names_list <- list()
for(i in 1:nrow(feat_index)){
  feat_names_list[[i]] <- feat_names[which(feat_index[i,] == 1)]
}
feat_comb <- length(feat_names_list)
```

**Output:** Predictive modeling feature combinations list, `data/feat_comb.rds`

```{r output-feat-comb, eval = F}
saveRDS(feat_comb, "data/feat_comb.rds")
```

There are ~9,000 feature combinations for 2-5 predictors.

### Join DOC and Features

```{r join-data, include = F, eval = F}
data <- doc %>% 
  left_join(feat, by = "IP_ID")
```

**Output:** Predictive modeling training dataset, `data/training-data.csv`

```{r output-training-data, eval = F}
write_csv(data, "data/training-data.csv")
```

## Part 2: Train Random Forest

```{r read-data, output = F, message = F}
train <- read_csv("data/training-data.csv")
feat_comb <- readRDS("data/feat_comb.rds")
```

```{r  set-seed, output = F, message = F}
set.seed(23)
IP_IDs <- train %>% 
  select(IP_ID) %>% pull
```

### Cluster Watersheds

```{r cluster-watersheds}
# site_loc <- read_csv("data/WtsData_For_RCN_DOCflux.csv") %>% 
#   group_by(IP_ID) %>% 
#   summarize(Lat = Poly_y[1],
#             Lon = Poly_x[1]) 
# 
# x <- site_loc$Lon
# y <- site_loc$Lat
# xy <- SpatialPointsDataFrame(matrix(c(x,y), ncol=2), data.frame(IP_ID=site_loc$IP_ID),
#                              proj4string = CRS("+proj=aea +ellps=GRS80 +datum=NAD83"))
# plot(xy)
# crs(xy)
# 
# xy <- spTransform(xy, CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")) 

# # calculate Euclidian distance 
# site_dist <- xy %>% distm()
# 
# # cluster
# site_clusters <- hclust(as.dist(site_dist), method ='complete')
# 
# # define 300 km threshold
# d <- 100000
# xy$fold <- cutree(site_clusters, h=d) 
# xy <- xy %>% as_tibble() %>% 
#   dplyr::select(IP_ID, fold)
# 
# # rejoin folds
# train <- train %>% 
#   left_join(xy, by = c("IP_ID")) %>%
#   dplyr::select(fold = fold, everything())

```


